---
title: "Supplementary Tables — Early Development of *Pocillopora verrucosa* Embryos and Larvae"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(DT)

# Helper: render a CSV as a searchable, paginated datatable.
# If the file does not exist, prints a note instead.
show_table <- function(path, caption) {
  if (!file.exists(path)) {
    cat("> **Note:** No significant results were found for this comparison;",
        "the table was not generated.\n")
    return(invisible(NULL))
  }
  read.csv(path, check.names = FALSE) |>
    dplyr::select(-any_of("X")) |>       # drop row-index column written by write.csv
    DT::datatable(
      caption    = caption,
      rownames   = FALSE,
      filter     = "top",
      extensions = "Buttons",
      options    = list(
        pageLength = 25,
        scrollX    = TRUE,
        dom        = "Bfrtip",
        buttons    = c("csv", "excel")
      )
    )
}
```

---

## Table S1. GO enrichment — all expressed genes, embryo stage (28°C)

Gene Ontology (GO) enrichment results (GOseq, Wallenius method) for all genes expressed
in embryos at 28°C. BH-adjusted p-values across all three ontologies (BP, MF, CC) are
shown. Generated in `scripts/1_Embryo_GeneExpression_28.Rmd`.

```{r tableS1}
show_table("figures/TableS1_Embryo_28_Enriched_GOTerms.csv",
           "Table S1. GO enrichment — expressed genes at embryo stage (28°C)")
```

---

## Table S2. GO enrichment — all expressed genes, larval stage (28°C)

Gene Ontology (GO) enrichment results (GOseq, Wallenius method) for all genes expressed
in larvae at 28°C. BH-adjusted p-values across all three ontologies (BP, MF, CC) are
shown. Generated in `scripts/2_Larval_GeneExpression_28.Rmd`.

```{r tableS2}
show_table("figures/TableS2_Larval_28_Enriched_GOTerms.csv",
           "Table S2. GO enrichment — expressed genes at larval stage (28°C)")
```

---

## Table S3. Annotated DEGs — embryo vs larval stage (28°C, BH-adjusted p < 0.01)

All differentially expressed genes (DEGs) between embryo and larval stages at 28°C,
with functional annotation. Positive log₂ fold change indicates higher expression in
larvae. Generated in `scripts/3_Larval_vs_Embryo_GeneExpression_28.Rmd`.

```{r tableS3}
show_table("figures/TableS3_DEG_01FDR_emb_vs_lar_28.csv",
           "Table S3. Annotated DEGs — embryo vs larval 28°C (BH adj p < 0.01)")
```

---

## Table S4. GO enrichment — embryo-upregulated DEGs (Figure 3A, BH-adjusted p < 0.01)

Biological Process (BP), Molecular Function (MF), and Cellular Component (CC) GO
enrichment results for genes significantly higher in embryos than larvae at 28°C.
Includes BH-adjusted p-values and gene ratio. Generated in
`scripts/3_Larval_vs_Embryo_GeneExpression_28.Rmd`.

```{r tableS4}
show_table("figures/TableS4_EMB_upregDEG_Enriched_GO_Terms.csv",
           "Table S4. GO enrichment — embryo-upregulated DEGs (BH adj p < 0.01)")
```

---

## Table S5. GO enrichment — larval-upregulated DEGs (Figure 3B, BH-adjusted p < 0.01)

Biological Process (BP), Molecular Function (MF), and Cellular Component (CC) GO
enrichment results for genes significantly higher in larvae than embryos at 28°C.
Includes BH-adjusted p-values and gene ratio. Generated in
`scripts/3_Larval_vs_Embryo_GeneExpression_28.Rmd`.

```{r tableS5}
show_table("figures/TableS5_LAR_upregDEG_Enriched_GO_Terms.csv",
           "Table S5. GO enrichment — larval-upregulated DEGs (BH adj p < 0.01)")
```

---

## Table S6. Annotated DEGs — 28°C vs 31°C larvae (BH-adjusted p < 0.05)

All differentially expressed genes between larvae reared at 28°C and 31°C, with
functional annotation. Positive log₂ fold change indicates higher expression at 31°C.
Generated in `scripts/4_Larval_GeneExpression_28_vs_31.Rmd`.

```{r tableS6}
show_table("figures/TableS6_28vs31_Larvae_Annotated_DEGS.csv",
           "Table S6. Annotated DEGs — 28°C vs 31°C larvae (BH adj p < 0.05)")
```

---

## Table S7. GO enrichment — 28°C-upregulated larvae (Figure 4, BH-adjusted p < 0.05)

GO enrichment results for genes significantly higher in larvae at 28°C relative to 31°C.
Generated in `scripts/4_Larval_GeneExpression_28_vs_31.Rmd`.

```{r tableS7}
show_table("figures/TableS7_L28_upregDEG_Enriched_GO_Terms.csv",
           "Table S7. GO enrichment — 28°C-upregulated larvae (BH adj p < 0.05)")
```

---

## Table S8. GO enrichment — 31°C-upregulated larvae (Figure 5, BH-adjusted p < 0.05)

GO enrichment results for genes significantly higher in larvae at 31°C relative to 28°C.
Generated in `scripts/4_Larval_GeneExpression_28_vs_31.Rmd`.

```{r tableS8}
show_table("figures/TableS8_L31_upregDEG_Enriched_GO_Terms.csv",
           "Table S8. GO enrichment — 31°C-upregulated larvae (BH adj p < 0.05)")
```
